---
title: "Population Genetics Assignment"
author: "31040462"
format: 
  html:
    embed-resources: true
---
### RUN ADMIXTURE
I run the following script in bash and obtained the .Q file.
```{bash}
#| eval: false

#!/bin/bash
# RUN ADMIXTURE on HapMap3 sample data

#----------Step 1: LD Pruning----------
# Use PLINK on the HapMap3 sample data

plink  --bfile hapmap3 \
       --indep-pairwise 50 10 0.1

# creates a new BED file with only pruned SNPs

plink  --bfile hapmap3 \
       --extract plink.prune.in \
       --make-bed \
       --out prunedData

#----------Step 2: Run ADMIXTURE ----------
# Run cross validation (CV) and ADMIXTURE on the pruned dataset
for K in 1 2 3 4 5; \
 do ./admixture --cv prunedData.bed $K | tee log${K}.out;
done

# view CV error for each K
grep -h CV log*.out
```

### CREATING ADMIXTURE PLOT
Load the required libraries
```{r}
#| message: false
# Load Libraries
library(tidyverse)
```

### Barchart showing ADMIXTURE
```{r}
# ========== Creating the barchart ==========

# Read in the .Q file with pruned data
tbl <- read.table("prunedData.3.Q")

# Create the barchart
barplot(t(as.matrix(tbl)),
        col = rainbow(3),
        xlab ="Individuals",
        ylab ="Ancestry",
        border = NA,
        main = "ADMIXTURE Results (K=3)")
```

### ADMIXTURE PLOT WITH GGPLOT2
```{r}
# ===================== READ IN FILES ===========================

# Read the .Q file for ancestry proportions
Q_file <- read.table("prunedData.3.Q")

# Read the .fam file to get individual IDs and their order

fam_file <- read.table("prunedData.fam")

# Read the tsv file for population metadata (population labels)
pop_file <- read.table("igsr_samples-2.tsv",
                       header = TRUE,
                       sep = "\t",
                       stringsAsFactors = FALSE)

# view the different files
head(Q_file, n = 2)
head(fam_file, n = 2)
head(pop_file, n = 2)

# =================== MERGE ALL INFORMATION ==================
# Create one dataframe containing info from the Q, fam and tsv files 

# Add sample IDs from fam file to Q file 
Q_file$SampleID <- fam_file$V2

colnames(Q_file)

# Identify the matching column in the TSV file
colnames(pop_file)
head(pop_file)
# sample IDs are in column Sample.name

# Merge the Q_file with pop_file
merged_data <- merge(Q_file,
                     pop_file,
                     by.x = "SampleID",
                     by.y = "Sample.name",
                     all.x = TRUE)

# view column names of the merged dataframe
colnames(merged_data)

# Rename ancestry columns for clarity
colnames(merged_data)[2:4] <- c("K1", "K2", "K3")
colnames(merged_data)

# =================== CONVERT TO LONG FORMAT ===================
# Convert the merged dataframe from wide format to long format

long_data <- merged_data |>
  pivot_longer(cols = starts_with ("K"),
               names_to = "Cluster",
               values_to = "Ancestry") |>
  arrange(Population.name, SampleID)

# Remove samples without population labels
long_data <- long_data |>
  filter(!is.na(Population.name),   
         !is.na(SampleID),
         !is.na(Ancestry))

# =================== CREATE THE PLOT ============================

ggplot(long_data,
       aes(x = SampleID,
           y = Ancestry,
           fill = Cluster)) +
  
  facet_grid(~ Population.name, scales = "free_x", space = "free_x", switch = "x") +
  
  geom_bar(stat = "identity",
           width = 1,
           color = NA,
           position = position_stack(reverse = TRUE)) +

  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.line.x = element_blank(),
        
        panel.border = element_blank(),
        strip.text.x = element_text(angle = 45, size = 10),
        strip.placement = "outside",
        strip.background = element_blank(),
        
        legend.position = "right",
        legend.title = element_blank(),
        panel.spacing = unit(0, "lines")
        ) +
  
  labs(title = "ADMIXTURE Plot (K=3) - HapMap3",
       x = "Individuals (Grouped by Population)",
       y = "Ancestry Proportion")

```

### EXPLANATION
The ADMIXTURE plot (K=3) shows the estimated ancestry proportions of individuals from the HapMap3 dataset, assuming three ancestral populations. Each vertical bar represents one individual, and the different colors show the proportion of ancestry from each inferred group. Individuals are grouped by their reported population such as CEPH and Yoruba.

The African Ancestry SW and Mexican Ancestry groups show clear evidence of admixture, where individuals possess segments from multiple clusters. This pattern reflects historical mixing between populations, resulting in more diverse genetic backgrounds.

